---
title: "Delivery_statistics"
output: html_document
date: "2026-05-21"
---
# Statistical Analysis

```{r}
# libraries
library(dplyr)
library(readr)
library(corrplot)
```

```{r}
data <- read_csv("delivery_clean.csv", col_names = TRUE)

data
```

## Description statistics

```{r}
summary(data)
```
## Weekend vs Weekday Earnings

- description (globally)
```{r}

data %>%
  summarise(
    avg_total = mean(total, na.rm = TRUE),
    median_total = median(total, na.rm = TRUE),
    sd_total = sd(total, na.rm = TRUE),
    avg_tip = mean(tip, na.rm = TRUE),
    avg_duration = mean(duration_minutes, na.rm = TRUE)
  )
```
- weekend vs working weekdays
```{r}
data %>%
  group_by(is_weekend) %>%
  summarise(
    avg_total = mean(total, na.rm = TRUE),
    avg_duration = mean(duration_minutes, na.rm = TRUE),
    avg_tip = mean(tip, na.rm = TRUE)
  )
```

- t-test
```{r}
t.test(
  total ~ is_weekend,
  data = data
)
```
- interpretation:

Weekend deliveries generated significantly different earnings compared to weekdays (p < 0.05).

## Weather vs Delivery Duration

- correlation Duration vs Wind speed
```{r}
cor(data$duration_minutes,
    data$wind,
    use = "complete.obs")
```
Wind speed showed a statistically significant positive relationship with delivery duration.

```{r}
numeric_df <- data %>%
  select(
    total,
    tip,
    duration_minutes,
    temp,
    wind,
    fall,
    factor
  )

cor_matrix <- cor(
  numeric_df,
  use = "complete.obs"
)

corrplot(cor_matrix)
```
- regression
```{r}
model <- lm(
  duration_minutes ~
    wind +
    temp +
    fall +
    factor +
    is_weekend,
  data = data
)
```

```{r}
summary(model)
```
Linear regression analysis indicated that temperature and rainfall were statistically significant predictors of delivery duration. Higher temperatures were associated with shorter delivery times, while rainfall unexpectedly showed a negative relationship with duration. Other variables, including wind speed, weekend status, and delivery multiplier, were not statistically significant predictors.

Despite statistical significance, the model explained only a small portion of delivery duration variability (R² ≈ 0.04), suggesting that additional operational factors such as traffic conditions and route characteristics likely play a major role.

## Time-of-Day Effects
- ANOVA
```{r}
anova_model <- aov(
  duration_minutes ~ factor(hour),
  data = data
)

```

```{r}
summary(anova_model)
```
- conclusions:

Despite visible variation across hours, ANOVA results indicated no statistically significant differences in average delivery duration.
It appears to depend more on operational and environmental factors than on time-of-day alone.

